Prediction of 90-day mortality after Total Hip Arthroplasty: a simplified and externally validated model based on observational registry data from Sweden, England and Wales

### 2020-05-23 reports/article.Rmd
Change ref style. Fix #26.

---
title: "Prediction of 90-day mortality after Total Hip Arthroplasty: a simplified and externally validated model based on observational registry data from Sweden, England and Wales"
author: Anne Garland`*`^1,2,3^, Erik Bülow`*`^2,4^, Erik Lenguerrand^5^, Ashley Blom^5,6^, J. Mark Wilkinson^7^, Adrian Sayers^5^, Ola Rolfson^2,4^, Nils P. Hailer^1,2^
date: '`r Sys.Date()`'
bibliography: H:/zotero_lib.bib
csl: the-lancet.csl
output: 
  bookdown::word_document2:
    reference_docx: article.dotx
    toc: false
    df_print: kable
---

@@

Shared decision making has evolved into an integral part of patient-physician interactions prior to surgical interventions, and the weighing of risks against benefits is central to this process. In terms of gained quality of life and cost-utility, total hip arthroplasty (THA) is an enormously successful procedure. 90-day mortality after THA performed due to osteoarthritis is low, ranging from 0.2 % to 0.6 % in large joint registry studies.[@Hunt2013; @Garland2017] The risk of short-term mortality is nevertheless an essential part of preoperative discussions between patients and surgeons.

Co-morbidity is associated with a shorter remaining life span, but the Charlson and Elixhauser co-morbidity indices poorly predict mortality after THA.[@Inacio2015; @Gordon2013; @Hofstede2016; @Garland2017; @Bulow2017; @Inacio2015] Additionally, these complex co-morbidity instruments are based on information from extensive in- or out-patient databases, each measure being defined by 1,178 and 1,516 individual International Classification of Diseases (ICD)-10 codes, respectively. Each co-morbidity index also exists in numerous versions, making interpretation and comparisons difficult.[@Brusselaers2017; @Walraven2009] The use of these indices is therefore limited to research settings but they are impossible to use under clinical circumstances.

Pre-operative co-morbidity data have been used to develop clinically applicable prediction models for early mortality and adverse events such as surgical site infection in the context of THA surgery.[@Bozic2013a; @Bozic2013b; @Harris2018; @Harris2018a; @Inacio2016; @Price2019; @Harris2019] The first prediction model based on a European cohort was the patient decision support tool for joint remplacement in the UK (https://jointcalc.shef.ac.uk/). However, it remains unclear to what level the depth of co-morbidity characterization influences the accuracy of  mortality prediction models.

Here we aimed to develop a parsimonious prediction model of 90-day postoperative mortality after THA with internal and external validation of discrimination and calibration, and to compare this to the accuracy of prediction models based on existing co-morbidity measures.[@Manning2016]

@@

We used data from the Swedish Hip Arthroplasty Register (SHAR) for model derivation and internal validation of discrimination and calibration. The best model was then validated externally on patients from England and Wales recorded in the National Joint Registry for England, Wales, Northern Ireland, the Isle of Man and the States of Guernsey (NJR). We focused on patients with cemented THA due to osteoarthritis. Cementation is the most common fixation technique used in Sweden, thus used as inclusion criteria to decrease heterogeneity. Only the last operated hip was accounted for in patients with bilateral THA, since those are more likely to resemble patients with unilateral THA (Figure \@ref(fig:flowchart)).[@Bulow2020]

@@

The patient cohort was identified from the SHAR. The inclusion period started 2008 since the American Society for Anesthesiologists (ASA) class and Body Mass Index (BMI) were systematically recorded from then on. The observation period ended 2015, since we had access to co-morbidity from the national patient register up to this point. Deterministic data linkage was performed by 10-digit identity numbers, assigned to all Swedish residents at birth or immigration.[@Ludvigsson2009] We used data from a variety of sources, as previously described.[@Cnudde2016] Age, sex, BMI, ASA class, type of hospital (university/county/rural/private) and year of surgery were collected from the SHAR, with a completeness of 96-98 %.[@Karrholm2018] Data on educational level (low = up to 9 years/middle = 10-12 years/high = at least 12 years) was recorded for more than 98 % of the population with 85 % accuracy in the longitudinal integration database for health insurance and labor market studies (LISA) from Statistics Sweden.[@Ludvigsson2019] Civil status (married/un-married/divorced/widow[er]), were also collected from LISA. The Swedish national patient register (NPR) was used to assess co-morbidity during the year preceding index surgery. This register contains all relevant diagnoses coded by ICD-10, as well as dates of admission and discharge for in- and outpatient episodes in all private and public hospitals. Completeness for NPR is above 99 % and 85-95 % of all diagnostic codes are valid.[@Ludvigsson2011] Death dates were linked from the national population register.

@@

Similar inclusion criteria were applied to the validation cohort from England and Wales. We excluded no patients based on missing information on BMI, educational level, or civil status however, since those variables were not used in the final model. The validation cohort was created by linking NJR data to the secondary uses service database from the National Health Service (NHS). This database includes co-morbidity recorded in the admitted patient care module of the hospital episode statistics (HES) database in England, and the patient episode database for Wales.[@Brophy2006; @Burns2012; @Sinha2013; @Simmonds2014; @Thorn2016] Out-patient deaths were linked from the office of national statistics. Data linkage used NHS-numbers, as well as patient name, age, sex and address.[@Sayers2016]

@@

Co-morbidity was defined by individual ICD-10 codes grouped into 17 categories according to Charlson [@Charlson1987; @Deyo1992; @Quan2005] and 31 categories according to Elixhauser.[@Elixhauser1998; @Quan2005] Some co-morbidities were identified in both indices, and some distinct co-morbidities were closely related (such as hypertension with and without complications, or abuse of either drugs or alcohol). We combined individual diagnostic groups to establish `r nrow(tab_categorization)` broader categories of co-morbidity (Table \@ref(tab:tabcategorization)). Conditions were merged according to clinical relevance as to be recognized in a patient-physician meeting without access to external register data.

@@

We used the Kaplan-Meier estimator to assess unadjusted mortality after cemented THA. Further analysis was based on logistic regression, since no loss to follow up occurred in the Swedish cohort within the first 90 days, and we assumed the same was true for patients from England and Wales. We used a modelling procedure with bootstrap ranking and a logistic least absolute shrinkage and selection operator (LASSO).[@Guo2015; @Baranowski2020] Numeric variables (age and BMI) were normalized before modelling to have mean = 0 and standard deviation = 1. Thus, the magnitude (absolute values) of their estimated coefficients would indicate variable importance on the same scale as the categorical variables. Co-morbidities recorded for at least one patient who died within 90 days, and one who did not, were included in the modelling process. 1,000 bootstrap samples were drawn from the observed data set.[@Austin2004] We used 10-fold cross validation for every bootstrap sample with a broad range of potential penalty values ($\lambda$:s) in a logistic LASSO model. We then only kept $\lambda$:s minimizing the mean cross-validated deviances in each sample. Those $\lambda$:s were used to estimate model coefficients for each potential predictor. The magnitude (absolute values) of those estimates were used as a measure of variable importance. Piece-wise linear regression was used to detect a breaking point where a significant drop in variable importance was observed. Potential predictors with variable importance above this breaking point were considered important and kept as model candidates. The whole process was repeated 100 times. Covariates that were selected at least once out of the 100 times were used in a main effects model of multivariable logistic regression without penalty, and without pre-normalization of numeric variables (main model). A reduced model with variables chosen at least 33 out of the 100 times was used as a simpler alternative for comparison. This model, without cancer as a predictor, was also evaluated separately, considering that medical indications for THA surgery may be different for patients with cancer compared to patients without malignancies. Univariable models with the ASA class, Charlson or Elixhauser co-morbidity indices were used for benchmarking, as well as a multivariable model including only age and sex. Each model including age was fitted three times, once with age as a main effect and twice with restricted cubic splines, either by two or three knots. Odds ratios for the final model were estimated with 95 % confidence intervals.[@Ripley2002]

@@

Each model was used to predict the probability of death within 90 days for patients from the SHAR (internal validation). Sensitivity and specificity were estimated to form receiver operating characteristic (ROC) curves and the area under those curves (AUC) were used as a measure of discriminative ability.[@Fawcett2006] Models with a lower 95 % confidence limit above 0.7, were considered good. Those intervals were based on percentiles from 2,000 non-parametric bootstrap samples. We used the bias-corrected Somers' $D_{xy}$ rank correlation based on 100 resamples to adjust for optimism.[@Miller1991] Calibration bands were made to graphically assess model calibration, comparing predicted probabilities and observed proportions.[@Nattino2016] The reduced model was then evaluated externally. An AUC with 95 % confidence interval was calculated for the model as-is. Re-calibration of the model intercept was then performed to account for different mortality rates in Sweden compared to England and Wales. An updated over-all slope was also calculated to account for country-specific treatment differences.[@Steyerberg2004] Calibration for this re-calibrated model was illustrated in the same calibration belt plot as for the internal calibration.

@@

Our model could be used in clinical practice, either by the formula above, or by accessing a simple web calculator online (`r shiny_url`). For example, a `r min(df$P_Age)` year old woman with ASA class I and none of the important co-morbidities would have a `r get_risk("baseline")` risk to die within 90 days of surgery. A woman, `r quantile(df$P_Age, .25)` years old, i.e. the first age quantile, would have an elevated risk of `r get_risk("healthy_female")`. A `r max(df$P_Age)` year old man, i.e. the maximum observed age, with ASA class III and cancer, would have a risk of `r get_risk("sickest_man")`. Note, however, that covariate patters with observed probabilities above 5 % were rare (`r mean(pred_probs > 5) * 100` %, n = `r sum(pred_probs > 5)`), and estimated mortality risks above 5 % are therefore subject to extrapolation. The observed proportions of deaths for patients with similar characteristics is likely lower (as indicated by Figure \@ref(fig:calibration)). Some risk calculators ignore this problem,[@Bozic2013a; @Harris2019] but we think this should be acknowledged.

Variables in our model were chosen based on statistical properties and not due to clinical relevance. Variables and estimated coefficients should therefore not be assigned any exact epidemiological and/or causal meaning per se.[@Breiman2001]

Age and sex are well-known predictors of remaining life span, but it is less obvious that this relation must be linear. We used restricted cubic splines to allow a more flexible relation, but found no difference in AUC compared to simpler main effect models. ASA class III has the largest estimated coefficient among all predictors, indicating large relative importance. This is clinically reasonable since a label of "severe systemic disease" should be based on a relevant patient assessment prior to surgery. ASA class is nevertheless known to have a high degree of internal variability.[@Haynes1995] It has previously been compared to the Charlson co-morbidity index, but not with respect to mortality after THA.[@Whitmore2014; @Kork2015] Patients with ASA class IV-VI were excluded since those categories describe severe disease, moribund and brain-dead individuals. We suspect that patients with such conditions might have been misclassified.

Obesity was not statistically significant by itself ($p =$ `r subset(brlasso_tbl_coefs, term == "Obesity", "p", drop = TRUE)`) but was still relevant as a predictor due to unobserved heterogeneity.[@Mood2010] We noted that the proportion of patients with BMI above 30, the WHO definition of obesity was much higher (`r round(mean(df$P_BMI > 30, na.rm = TRUE) * 100)` %) in the Swedish cohort than the proportion of patients with a diagnosis code for obesity (ICD-10 = E66) recorded in the national patient register (`r round(mean(df$c_obesity) * 100, 1)` %). We believe this might be due to the fact that BMI is systematically recorded in the SHAR, but the diagnosis of obesity might not have been made during previous contacts with the healthcare system.

@@

A strength of this study is its nationwide design with data from both Sweden, England and Wales. The Swedish registers are valid with low proportions of missing data.[@Soderman2000; @Karrholm2018; @Soderman2001; @Ludvigsson2011] Some concerns have been raised regarding validity of the hospital episodes statistics (HES) database from England and Wales however.[@Spencer2012; @Williams2002; @Brennan2012] A systematic review found that the overall median diagnostic accuracy (comparing ICD-codes from HES to individual case notes) was 80 %.[@Burns2012] Both in- and outpatient co-morbidity data were available for Sweden, but only in-patient data for England and Wales. We did not have access to any data from primary care or nursing homes in any of the counties.

The linkage procedure between NJR and HES has been previously described,[@Smith2012] where it was noted that privately funded patients were not included in HES. 17 % of the patients had missing personal data or did not allow linkage, and 6 % were not found in HES although their data were available from the NJR.

@@

We investigated whether the effect of age on mortality was non-linear using restricted cubic splines. A similar approach might be relevant for BMI, which we only modelled as a linear main effect. A U-shaped association between mortality and BMI categorized as underweight (below 18.5 kg/m^2^), normal (18.5 to 25 kg/m^2^), or obese (above 25 kg/m^2^) has been found for example in England and Wales.[@Mouchti2018] Our recordings of BMI might be imprecise however.[@Stommel2009] Some hospitals might report actual measurements but it is unknown to what extent those data are self-estimated by the patients or estimated by health care personnel. Therefore, we aimed to avoid overfitting BMI based on too elaborate transformations.

>>>>

---
title: "Prediction of 90-day mortality after Total Hip Arthroplasty: a simplified and externally validated model based on observational registry data from Sweden, England and Wales"
author: Anne Garland`*`^1,2,3^, Erik Bülow`*`^2,4^, Erik Lenguerrand^5^, Ashley Blom^5,6^, J. Mark Wilkinson^7^, Adrian Sayers^5^, Ola Rolfson^2,4^, Nils P. Hailer^1,2^
date: '`r Sys.Date()`'
bibliography: H:/zotero_lib.bib
csl: bmc-medicine.csl
output: 
  bookdown::word_document2:
    reference_docx: article.dotx
    toc: false
    df_print: kable
---

@@

Shared decision making has evolved into an integral part of patient-physician interactions prior to surgical interventions, and the weighing of risks against benefits is central to this process. In terms of gained quality of life and cost-utility, total hip arthroplasty (THA) is an enormously successful procedure. 90-day mortality after THA performed due to osteoarthritis is low, ranging from 0.2 % to 0.6 % in large joint registry studies [@Hunt2013; @Garland2017]. The risk of short-term mortality is nevertheless an essential part of preoperative discussions between patients and surgeons.

Co-morbidity is associated with a shorter remaining life span, but the Charlson and Elixhauser co-morbidity indices poorly predict mortality after THA [@Inacio2015; @Gordon2013; @Hofstede2016; @Garland2017; @Bulow2017]. Additionally, these complex co-morbidity instruments are based on information from extensive in- or out-patient databases, each measure being defined by 1,178 and 1,516 individual International Classification of Diseases (ICD)-10 codes, respectively. Each co-morbidity index also exists in numerous versions, making interpretation and comparisons difficult [@Brusselaers2017; @Walraven2009]. The use of these indices is therefore limited to research settings but they are impossible to use under clinical circumstances.

Pre-operative co-morbidity data have been used to develop clinically applicable prediction models for early mortality and adverse events such as surgical site infection in the context of THA surgery [@Bozic2013a; @Bozic2013b; @Harris2018; @Harris2018a; @Inacio2016; @Price2019; @Harris2019]. The first prediction model based on a European cohort was the patient decision support tool for joint remplacement in the UK (https://jointcalc.shef.ac.uk/). However, it remains unclear to what level the depth of co-morbidity characterization influences the accuracy of  mortality prediction models.

Here we aimed to develop a parsimonious prediction model of 90-day postoperative mortality after THA with internal and external validation of discrimination and calibration, and to compare this to the accuracy of prediction models based on existing co-morbidity measures [@Manning2016].

@@

We used data from the Swedish Hip Arthroplasty Register (SHAR) for model derivation and internal validation of discrimination and calibration. The best model was then validated externally on patients from England and Wales recorded in the National Joint Registry for England, Wales, Northern Ireland, the Isle of Man and the States of Guernsey (NJR). We focused on patients with cemented THA due to osteoarthritis. Cementation is the most common fixation technique used in Sweden, thus used as inclusion criteria to decrease heterogeneity. Only the last operated hip was accounted for in patients with bilateral THA, since those are more likely to resemble patients with unilateral THA (Figure \@ref(fig:flowchart)) [@Bulow2020].

@@

The patient cohort was identified from the SHAR. The inclusion period started 2008 since the American Society for Anesthesiologists (ASA) class and Body Mass Index (BMI) were systematically recorded from then on. The observation period ended 2015, since we had access to co-morbidity from the national patient register up to this point. Deterministic data linkage was performed by 10-digit identity numbers, assigned to all Swedish residents at birth or immigration [@Ludvigsson2009]. We used data from a variety of sources, as previously described [@Cnudde2016]. Age, sex, BMI, ASA class, type of hospital (university/county/rural/private) and year of surgery were collected from the SHAR, with a completeness of 96-98 % [@Karrholm2018]. Data on educational level (low = up to 9 years/middle = 10-12 years/high = at least 12 years) was recorded for more than 98 % of the population with 85 % accuracy in the longitudinal integration database for health insurance and labor market studies (LISA) from Statistics Sweden [@Ludvigsson2019]. Civil status (married/un-married/divorced/widow[er]), were also collected from LISA. The Swedish national patient register (NPR) was used to assess co-morbidity during the year preceding index surgery. This register contains all relevant diagnoses coded by ICD-10, as well as dates of admission and discharge for in- and outpatient episodes in all private and public hospitals. Completeness for NPR is above 99 % and 85-95 % of all diagnostic codes are valid [@Ludvigsson2011]. Death dates were linked from the national population register.

@@

Similar inclusion criteria were applied to the validation cohort from England and Wales. We excluded no patients based on missing information on BMI, educational level, or civil status however, since those variables were not used in the final model. The validation cohort was created by linking NJR data to the secondary uses service database from the National Health Service (NHS). This database includes co-morbidity recorded in the admitted patient care module of the hospital episode statistics (HES) database in England, and the patient episode database for Wales [@Brophy2006; @Burns2012; @Sinha2013; @Simmonds2014; @Thorn2016]. Out-patient deaths were linked from the office of national statistics. Data linkage used NHS-numbers, as well as patient name, age, sex and address [@Sayers2016].

@@

Co-morbidity was defined by individual ICD-10 codes grouped into 17 categories according to Charlson [@Charlson1987; @Deyo1992; @Quan2005] and 31 categories according to Elixhauser [@Elixhauser1998; @Quan2005]. Some co-morbidities were identified in both indices, and some distinct co-morbidities were closely related (such as hypertension with and without complications, or abuse of either drugs or alcohol). We combined individual diagnostic groups to establish `r nrow(tab_categorization)` broader categories of co-morbidity (Table \@ref(tab:tabcategorization)). Conditions were merged according to clinical relevance as to be recognized in a patient-physician meeting without access to external register data.

@@

We used the Kaplan-Meier estimator to assess unadjusted mortality after cemented THA. Further analysis was based on logistic regression, since no loss to follow up occurred in the Swedish cohort within the first 90 days, and we assumed the same was true for patients from England and Wales. We used a modelling procedure with bootstrap ranking and a logistic least absolute shrinkage and selection operator (LASSO) [@Guo2015; @Baranowski2020]. Numeric variables (age and BMI) were normalized before modelling to have mean = 0 and standard deviation = 1. Thus, the magnitude (absolute values) of their estimated coefficients would indicate variable importance on the same scale as the categorical variables. Co-morbidities recorded for at least one patient who died within 90 days, and one who did not, were included in the modelling process. 1,000 bootstrap samples were drawn from the observed data set [@Austin2004]. We used 10-fold cross validation for every bootstrap sample with a broad range of potential penalty values ($\lambda$:s) in a logistic LASSO model. We then only kept $\lambda$:s minimizing the mean cross-validated deviances in each sample. Those $\lambda$:s were used to estimate model coefficients for each potential predictor. The magnitude (absolute values) of those estimates were used as a measure of variable importance. Piece-wise linear regression was used to detect a breaking point where a significant drop in variable importance was observed. Potential predictors with variable importance above this breaking point were considered important and kept as model candidates. The whole process was repeated 100 times. Covariates that were selected at least once out of the 100 times were used in a main effects model of multivariable logistic regression without penalty, and without pre-normalization of numeric variables (main model). A reduced model with variables chosen at least 33 out of the 100 times was used as a simpler alternative for comparison. This model, without cancer as a predictor, was also evaluated separately, considering that medical indications for THA surgery may be different for patients with cancer compared to patients without malignancies. Univariable models with the ASA class, Charlson or Elixhauser co-morbidity indices were used for benchmarking, as well as a multivariable model including only age and sex. Each model including age was fitted three times, once with age as a main effect and twice with restricted cubic splines, either by two or three knots. Odds ratios for the final model were estimated with 95 % confidence intervals [@Ripley2002].

@@

Each model was used to predict the probability of death within 90 days for patients from the SHAR (internal validation). Sensitivity and specificity were estimated to form receiver operating characteristic (ROC) curves and the area under those curves (AUC) were used as a measure of discriminative ability [@Fawcett2006]. Models with a lower 95 % confidence limit above 0.7, were considered good. Those intervals were based on percentiles from 2,000 non-parametric bootstrap samples. We used the bias-corrected Somers' $D_{xy}$ rank correlation based on 100 resamples to adjust for optimism [@Miller1991]. Calibration bands were made to graphically assess model calibration, comparing predicted probabilities and observed proportions [@Nattino2016]. The reduced model was then evaluated externally. An AUC with 95 % confidence interval was calculated for the model as-is. Re-calibration of the model intercept was then performed to account for different mortality rates in Sweden compared to England and Wales. An updated over-all slope was also calculated to account for country-specific treatment differences [@Steyerberg2004]. Calibration for this re-calibrated model was illustrated in the same calibration belt plot as for the internal calibration.

@@

Our model could be used in clinical practice, either by the formula above, or by accessing a simple web calculator online (`r shiny_url`). For example, a `r min(df$P_Age)` year old woman with ASA class I and none of the important co-morbidities would have a `r get_risk("baseline")` risk to die within 90 days of surgery. A woman, `r quantile(df$P_Age, .25)` years old, i.e. the first age quantile, would have an elevated risk of `r get_risk("healthy_female")`. A `r max(df$P_Age)` year old man, i.e. the maximum observed age, with ASA class III and cancer, would have a risk of `r get_risk("sickest_man")`. Note, however, that covariate patters with observed probabilities above 5 % were rare (`r mean(pred_probs > 5) * 100` %, n = `r sum(pred_probs > 5)`), and estimated mortality risks above 5 % are therefore subject to extrapolation. The observed proportions of deaths for patients with similar characteristics is likely lower (as indicated by Figure \@ref(fig:calibration)). Some risk calculators ignore this problem [@Bozic2013a; @Harris2019], but we think this should be acknowledged.

Variables in our model were chosen based on statistical properties and not due to clinical relevance. Variables and estimated coefficients should therefore not be assigned any exact epidemiological and/or causal meaning per se [@Breiman2001].

Age and sex are well-known predictors of remaining life span, but it is less obvious that this relation must be linear. We used restricted cubic splines to allow a more flexible relation, but found no difference in AUC compared to simpler main effect models. ASA class III has the largest estimated coefficient among all predictors, indicating large relative importance. This is clinically reasonable since a label of "severe systemic disease" should be based on a relevant patient assessment prior to surgery. ASA class is nevertheless known to have a high degree of internal variability [@Haynes1995]. It has previously been compared to the Charlson co-morbidity index, but not with respect to mortality after THA [@Whitmore2014; @Kork2015]. Patients with ASA class IV-VI were excluded since those categories describe severe disease, moribund and brain-dead individuals. We suspect that patients with such conditions might have been misclassified.

Obesity was not statistically significant by itself ($p =$ `r subset(brlasso_tbl_coefs, term == "Obesity", "p", drop = TRUE)`) but was still relevant as a predictor due to unobserved heterogeneity [@Mood2010]. We noted that the proportion of patients with BMI above 30, the WHO definition of obesity was much higher (`r round(mean(df$P_BMI > 30, na.rm = TRUE) * 100)` %) in the Swedish cohort than the proportion of patients with a diagnosis code for obesity (ICD-10 = E66) recorded in the national patient register (`r round(mean(df$c_obesity) * 100, 1)` %). We believe this might be due to the fact that BMI is systematically recorded in the SHAR, but the diagnosis of obesity might not have been made during previous contacts with the healthcare system.

@@

A strength of this study is its nationwide design with data from both Sweden, England and Wales. The Swedish registers are valid with low proportions of missing data [@Soderman2000; @Karrholm2018; @Soderman2001; @Ludvigsson2011]. Some concerns have been raised regarding validity of the hospital episodes statistics (HES) database from England and Wales however [@Spencer2012; @Williams2002; @Brennan2012]. A systematic review found that the overall median diagnostic accuracy (comparing ICD-codes from HES to individual case notes) was 80 % [@Burns2012]. Both in- and outpatient co-morbidity data were available for Sweden, but only in-patient data for England and Wales. We did not have access to any data from primary care or nursing homes in any of the counties.

The linkage procedure between NJR and HES has been previously described [@Smith2012], where it was noted that privately funded patients were not included in HES. 17 % of the patients had missing personal data or did not allow linkage, and 6 % were not found in HES although their data were available from the NJR.

@@

We investigated whether the effect of age on mortality was non-linear using restricted cubic splines. A similar approach might be relevant for BMI, which we only modelled as a linear main effect. A U-shaped association between mortality and BMI categorized as underweight (below 18.5 kg/m^2^), normal (18.5 to 25 kg/m^2^), or obese (above 25 kg/m^2^) has been found for example in England and Wales [@Mouchti2018]. Our recordings of BMI might be imprecise however [@Stommel2009]. Some hospitals might report actual measurements but it is unknown to what extent those data are self-estimated by the patients or estimated by health care personnel. Therefore, we aimed to avoid overfitting BMI based on too elaborate transformations.
